---
title: "ORIEN_CIBERSOTX_NSCLC2025"
author: "Dipankor Dhrubo"
date: "2025-10-07"
output: html_document
---

```{r setup, include=FALSE}
library("tidyverse")
library(survival)
library(ggsurvfit)
library(survminer)
```


```{r}
# Loading data

lung_expression<- read_tsv("/fs/ess/PAS1695/projects/exotho/data/microbiomeRIG/THO_TPM.txt")
Metadata<- readRDS(file= "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/NSCLC_metadata.rds")
Gene_symbol<- read_csv("/fs/ess/PAS1695/projects/exotho/data/microbiomeRIG/key_gene-id-to-symbol.csv")

##
lung_expression<- lung_expression %>% left_join(Gene_symbol, by="gene_id") %>% select(-gene_id)
# 1) Keep rows with a gene symbol
keep <- !is.na(lung_expression$gene_symbol) & lung_expression$gene_symbol != ""

# 2) Build a numeric matrix of expression only
expr <- lung_expression[keep, setdiff(names(lung_expression), "gene_symbol"), drop = FALSE]
# Force to numeric in case some columns are characters
expr[] <- lapply(expr, function(x) suppressWarnings(as.numeric(x)))
mat <- as.matrix(expr)

# 3) Collapse duplicates by SUM using rowsum
g <- lung_expression$gene_symbol[keep]
lung_expression2 <- rowsum(mat, group = g)  # genes x samples matrix

# 4) If you need a data.frame:
lung_expression2 <- as.data.frame(lung_expression2)

colnames(lung_expression2) <- sub("\\.genes\\.results$", "", colnames(lung_expression2))
samples<- unique(Metadata$sample)

lung_expression2<- lung_expression2 %>% select(all_of(samples))


#### CibersortX Immune cell analysis

gene_matrix1 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1:201)


write.table(gene_matrix1, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix1.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix2 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,202:401)

write.table(gene_matrix2, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix2.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix3 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,402:601)

write.table(gene_matrix3, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix3.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix4 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,602:801)

write.table(gene_matrix4, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix4.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix5 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,802:946)

write.table(gene_matrix5, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix5.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix6 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,947:1334)

write.table(gene_matrix6, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix6.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix7 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,1335:1536)

write.table(gene_matrix7, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix7.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)


gene_matrix8 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,1537:1737)

write.table(gene_matrix8, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix8.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

gene_matrix9 <- 
  lung_expression2 %>% 
  rownames_to_column("GeneSymbol") %>% 
  select(1,1738:1844)

write.table(gene_matrix9, "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/cibersortx_input_matrix9.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)


```

```{r}
#calling in outputs

matrix1 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input1_Results.csv")
matrix2 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input2_Results.csv")
matrix3 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input3_Results.csv")
matrix4 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input4_Results.csv")
matrix5 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input5_Results.csv")
matrix6 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input6_Results.csv")
matrix7 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input7_Results.csv")
matrix8 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input8_Results.csv")
matrix9 <- read_csv("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/CIBERSORTx_input9_Results.csv")
immune_matrix <- rbind(matrix1, matrix2, matrix3, matrix4, matrix5, matrix6,matrix7,matrix8,matrix9)

saveRDS(object = immune_matrix,
        file = "/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/immune_deconvolveORIEN.RDS")




immune_deconvolution <- readRDS("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Data/immune_deconvolveORIEN.RDS") %>% 
  rename_with(~ "sample", .cols = matches("Mixture")) %>% 
  select(-c("P-value", "Correlation", "RMSE")) %>% 
  pivot_longer(
    cols = -sample,                
    names_to = "ImmuneCell",       
    values_to = "Abundance"        
  )


## Prepping for the analysis

joindf <- Metadata %>% select(sample,CAlevel) %>% unique()

plotdf <- immune_deconvolution %>% 
  left_join(joindf, by = "sample")


wilcox_results <- plotdf %>%
  group_by(ImmuneCell) %>%
  summarise(
    p_value = wilcox.test(Abundance ~ CAlevel)$p.value,
    .groups = "drop"
  ) 

sig_cells <- wilcox_results %>%
  filter(p_value < 0.05) %>% filter(! ImmuneCell == "T cells regulatory (Tregs)")
```


```{r}

### Plot visualization

ggplot(plotdf[plotdf$ImmuneCell %in% sig_cells$ImmuneCell,], 
       aes(x = CAlevel, y = Abundance, fill = CAlevel)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  facet_wrap(~ ImmuneCell, scales = "free_y", ncol = 4) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +  # Add p-values
  scale_fill_manual(values = c("Low" = "#00BFC4", "High" = "#F8766D")) +
  theme_bw(base_size = 12) +
  labs(title = "Immune Cell Fractions by Candida Level",
       x = "Candida Level", y = "Fraction") +
  theme(strip.text = element_text(face = "bold", size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("/users/PAS1695/dipankor99/Github/digital-twins/exploratory/Figures/Immune_cells_NSCLC.png", 
       width = 13, height = 8)


```


## Result Interpretation: 

NK cells kill stressed/“non-self” targets without prior priming and shape early anti-tumor immunity (e.g., by secreting IFN-γ and licensing DCs).

CD8⁺ T cells are the main cytotoxic lymphocytes that clear tumor cells after antigen recognition; their abundance/activation strongly correlates with better prognosis and response to IO.

So the observed drop in activated NK and CD8 T cells in Candida-High tumors suggests a shift toward a less cytotoxic, less surveillance-competent microenvironment. That fits your other signals:

M0 macrophages ↑ and mast cells resting ↑ → more immature/myeloid-skewed, potentially immunosuppressive milieu.

CD4 memory activated ↑ with CD8 ↓ → “inflamed but ineffective” pattern seen in exhausted/immune-excluded tumors.

Mechanistically, Candida (and hypoxia it may reinforce) could drive IL-6/STAT3, TNF/NF-κB, and metabolic competition → T-cell/NK dysfunction, poor antigen presentation, and resistance to chemo/RT/IO.
